Categorical and gradient aspects of wordlikeness\thanks{Thanks to Gene Buckley, Charles Yang, and audiences at NELS 43 for helpful comments.}

### 2013-01-17 3.tex
Pre-"revisions", with lots of material in the TODO

\noindent
However, subsequent sound changes \citep[e.g.,][]{Baldi1994,Safarewicz1932}, particularly the degemination of Old Latin \emph{ss} before diphthongs and long monophthongs (e.g., \emph{caussa} $>$ \emph{causa} `cause'), introduce numerous exceptions to \textsc{Rhotacism}.
In Classical Latin, intervocalic \emph{s} is found root-internally (\emph{asellus} `donkey', \emph{casa} `hut'), in environments derived by inflectional suffixes (\emph{uās}-\emph{uāsis} `vase', \emph{uisēre} `to view'), prefixation (\emph{dēsecāre} `to cut off'; cf.~\emph{dē} `from', \emph{secāre} `to cut'), compounding (\emph{olusātrum} `parsnip'; cf.~\emph{olus} `vegetable', \emph{ātrum} `black'), and denominal adjective formation (\emph{uentōsus} `windy'; cf.~\emph{uentus} `wind'), and is tolerated in nativized loanwords from Celtic (\emph{omāsum} `tripe'), Germanic (\emph{glaesum} `amber'), and Greek (\emph{basis} `pedestal').
These facts lead Saussure to conclude that \textsc{Rhotacism} is no longer ``inhérente à la nature de la langue'' (202).
While some of the apparent exceptions may be the result of opaque phonological application \citep{Heslin1987}---an explanation not yet available in Saussure's time---any formulation of rhotacization will admit nearly as many lexical exceptions as there are roots exhibiting \emph{s}-\emph{r} alternations \citep{GormanInPressc}.
Any synchronic account of the underrepresentation of intervocalic \emph{s} must confront and the unproductive nature (as indicated by the accumulation of exceptions) of the proximate explanation for this tendency.

@@

\citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule \emph{R} must be marked [$+$\emph{R}] by convention. 
A root or affix which fails to undergo \emph{R} despite otherwise matching the structural description is simply said to be marked [$-$\emph{R}]. 
In other words, no representation is never truly an exception; rather, some underlying representations have non-default features which do not match the extended structural description of $R$, which requires that the target be [$+R$]. 
If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the trigger, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.

@@

\citeauthor{A74} also notes that roots which fail to undergo suffix harmony, like (\ref{turkexcept}b), may themselves be harmonic. He takes to be evidence for the necessity of duplication:

\begin{quote}
\ldots{}there are words which are exceptions to harmony across boundaries (e.g., \emph{kabahat} `fault', \emph{kabahattı} `his fault') but which are perfectly regular internally. Since the morpheme structure condition and the phonological rule in this case have distinct classes of exceptions, it is clear that they cannot be identified. \citep[289]{A74}
\end{quote}

@@

\noindent
It is unclear whether the cluster-splitting vowel is deleted in the non-native variant or epenthesized in the fast speech variant. 
    Under either analysis, there is no ready explanation for the tendency of the cluster-splitting vowel to have the backness features of following consonants; if anything, one might expect it to determine the backness features of following consonants. 
    All that can be said with certainty is that the adaptation of non-native onset clusters appears to proceed in such a fashion so that the lexical items in question are [$-$\textsc{Backness Harmony}].

@@

\noindent 
In the full specification analysis adopted here, reharmonization is the result of \textsc{Backness Harmony} applying within the reduplicant. 
    On the other hand, the lack of reharmonization in the reduplicants of disharmonic roots suggests that the [$-$\textsc{Backness Harmony}] exception feature is copied under reduplication.\footnote{
    There is reason to suspect that reduplicants bear the lexical diacritics of the roots they are derived from.
    In Kinande, verb reduplication, 
    Kinande reduplication, documented by \citet{Mutaka1990} and further discussed by \citet{Downing2000}, requires a bisyllabic reduplicant. 
    This clearly has synchronic force in the gramar, since reduplicated monosyllabic roots contain three copies of the root so as to satisfy this requirement, and reduplicated forms of many trisyllabic verbs are ineffable. 
    However, a few trisyllabic verbs exceptionally show full reduplication.
    This is a lexical property, but presence or absense of this exceptionalty feature is only apparent in the form of the reduplicant.}
    Under this definition, root harmony is not static, and so this is not an exception to the claim that static phonotactic generalizations are not preserved by operations like reduplication and truncation \citep{Silverman2000}.

A number of studies have investigated the role of harmony in word-spotting tasks, though to mimic auditory word recognition and segmentation in natural settings. 
    Many of these studies have been carried out in Finnish, which has a vowel harmony system similar to that of Turkish. 
    \citet{Suomi1997} and \citet{Vroomen1998} task Finnish speakers with identifying harmonic disyllables in an auditory stream. 
    When the syllable preceding the disyllabic target has a different backness specification than the target, recognition of the target is facilitated. Presumably, disharmony facilitates the recognition of word boundaries. 
    \citet{Kabak2010} find that Turkish \textsc{Backness Harmony} has a similar effect: Turkish speakers are quicker and more accurate at the task of spotting the nonce target word \emph{pavo} when preceded by a disharmonic juncture (e.g., \emph{gölü-PAVO}) than when preceded by a harmonic juncture (e.g., \emph{golu-PAVO}). 
    \citeauthor{Kabak2010} report that this effect does not obtain for speakers of French, a language which lacks vowel harmony. 
    This implies that Turkish speakers have internalized the tendency of harmonic sequences to be root-internal and of disharmonic transitions to cross word boundaries.

@@

To test for associations between the process (more specifically, the constraint that it imposes on roots) and type frequency in this database, each root was sorted into a $2 \times 2$ contingency table; the contents of each cell are specific to the process in question. The counts in this table are not expected to add up to 9,601, since many roots neither could exemplify nor violate the process in question; for instance, monosyllabic words are irrelevant to root harmony. The Fisher exact test is used to compute a $p$-value representing the probability of the observed patterns arising under the null hypothesis that there is no association between the constraint and type frequency.

@@

\textsc{Backness Harmony} is exemplified in the lexicon insofar as there is an positive association between the backness of vowels in all adjacent syllables. Any disagreement on the backness specification of vowels in adjacent syllables are counted solely as exceptions, even if other vowel transitions in the root are harmonic; further assumptions are necessary to determine whether a root which has one disharmonic transition may in any sense ``obey'' harmony elsewhere.
%This follows from \emph{SPE} theory of lexical exceptionality (see also \citealt{Gouskova2012}), in which phonological exceptionality is a property of underlying representations, not individual segments. 
To construct the contingency table, roots are binned according to the backness specification of the first nucleus, and that of following nuclei. For example, the first two syllables of \emph{adalet} `justice' are harmonic, but it is coded as disharmonic because there is a \emph{a\ldots{}e} transition later in the word. The resulting counts are shown in Table \ref{bhs}. 61\% of roots conform to \textsc{Backness Harmony}, and the interaction between the backness of the first and of the subsequent vowels
%predicted by root-internal \textsc{Backness Harmony} 
is significant.

@@

\citet{Zimmer1969} administers two variants of the paired nonce word rating task. The first used 23 native adult speakers who were permitted to select either nonce word as more like Turkish, or to indicate `no preference'. For the purposes below, `no preference' results are of. The second experiment used 32 native adults, none of whom appeared in the preceding study, and used a forced binary choice.

>>>>

\noindent
However, subsequent sound changes \citep[e.g.,][]{Baldi1994,Safarewicz1932}, particularly the degemination of Old Latin \emph{ss} after diphthongs and long monophthongs (e.g., \emph{caussa} $>$ \emph{causa} `cause'), introduce numerous exceptions to \textsc{Rhotacism}.
In Classical Latin, intervocalic \emph{s} is found root-internally (\emph{asellus} `donkey', \emph{casa} `hut'), in environments derived by inflectional suffixes (\emph{uās}-\emph{uāsis} `vase', \emph{uisēre} `to view'), prefixation (\emph{dēsecāre} `to cut off'; cf.~\emph{dē} `from', \emph{secāre} `to cut'), compounding (\emph{olusātrum} `parsnip'; cf.~\emph{olus} `vegetable', \emph{ātrum} `black'), and denominal adjective formation (\emph{uentōsus} `windy'; cf.~\emph{uentus} `wind'), and is tolerated in nativized loanwords from Celtic (\emph{omāsum} `tripe'), Germanic (\emph{glaesum} `amber'), and Greek (\emph{basis} `pedestal').
These facts lead Saussure to conclude that \textsc{Rhotacism} is no longer ``inhérente à la nature de la langue'' (202).
While some of the apparent exceptions may be the result of opaque phonological application \citep{Heslin1987}---an explanation not yet available in Saussure's time---any formulation of rhotacization will admit nearly as many lexical exceptions as there are roots exhibiting \emph{s}-\emph{r} alternations \citep{GormanInPressc}.
Any synchronic account of the underrepresentation of intervocalic \emph{s} must confront the unproductive nature (as indicated by the accumulation of exceptions) of the proximate explanation for this tendency.

% FIXME mention Fudge's analysis of short-a before voiceless fricatives
% \citet{Trager1930,Trager1934,Trager1940},
% \citet{Cohen1970}
% \citet{Ferguson1975}
% \citet{PLC1} 429f.
% \citet{Fudge1987}
% \citet{AOE} I.203f., 515f. for an attempt to write a rule
% \citet{Payne1976,Payne1980} who finds that whereas speakers from out of state acquire 5 vocalic sound changes, only one out of 34 interviewees acquired the core pattern
% Fudge(!)

@@

\citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule \emph{R} must be marked [$+$\emph{R}] by convention. 
A root or affix which fails to undergo \emph{R} despite otherwise matching the structural description is simply said to be marked [$-$\emph{R}]. 
In other words, no representation is ever truly an exception; rather, some underlying representations have non-default features which do not match the extended structural description of $R$, which requires that the target be [$+R$]. 
If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the trigger, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.

@@

\citeauthor{A74} also notes that roots which fail to undergo suffix harmony, like (\ref{turkexcept}b), may themselves be harmonic. He takes this to be evidence for the necessity of duplication:

\begin{quote}
\ldots{}there are words which are exceptions to harmony across boundaries (e.g., \emph{kabahat} `fault', \emph{kabahatti} `his fault') but which are perfectly regular internally. Since the morpheme structure condition and the phonological rule in this case have distinct classes of exceptions, it is clear that they cannot be identified. \citep[289]{A74}
\end{quote}

@@

\noindent
It is unclear whether the cluster-splitting vowel is deleted in the non-native variant or epenthesized in the fast speech variant. 
    Under either analysis, there is no ready explanation for the tendency of the cluster-splitting vowel to have the backness features of following vowels; if anything, one might expect it to determine the backness features of following vowels. 
    All that can be said with certainty is that the adaptation of non-native onset clusters appears to proceed in such a fashion so that the lexical items in question are [$+$\textsc{Backness Harmony}].

@@

\noindent 
In the full specification analysis adopted here, reharmonization is the result of \textsc{Backness Harmony} applying within the reduplicant. 
    On the other hand, the lack of reharmonization in the reduplicants of disharmonic roots suggests that the [$-$\textsc{Backness Harmony}] exception feature is copied under reduplication.\footnote{
    There is reason to suspect that reduplicants bear the lexical diacritics of the roots they are derived from.
    In Kinande, verb reduplication, documented by \citet{Mutaka1990} and further discussed by \citet{Downing2000}, requires a bisyllabic reduplicant. 
    This clearly has synchronic force in the gramar, since reduplicated monosyllabic roots contain three copies of the root so as to satisfy this requirement, and reduplicated forms of many trisyllabic verbs are ineffable. 
    However, a few trisyllabic verbs exceptionally show full reduplication.
    This is a lexical property, but presence or absense of this exceptionalty feature is only apparent in the form of the reduplicant.}
    Under this definition, root harmony is not static, and so this is not an exception to the claim that static phonotactic generalizations are not preserved by operations like reduplication and truncation \citep{Silverman2000}.

A number of studies have investigated the role of harmony in word-spotting tasks, thought to mimic auditory word recognition and segmentation in natural settings. 
    Many of these studies have been carried out in Finnish, which has a vowel harmony system similar to that of Turkish. 
    \citet{Suomi1997} and \citet{Vroomen1998} task Finnish speakers with identifying harmonic disyllables in an auditory stream. 
    When the syllable preceding the disyllabic target has a different backness specification than the target, recognition of the target is facilitated. Presumably, disharmony facilitates the recognition of word boundaries. 
    \citet{Kabak2010} find that Turkish \textsc{Backness Harmony} has a similar effect: Turkish speakers are quicker and more accurate at the task of spotting the nonce target word \emph{pavo} when preceded by a disharmonic juncture (e.g., \emph{gölü-PAVO}) than when preceded by a harmonic juncture (e.g., \emph{golu-PAVO}). 
    \citeauthor{Kabak2010} report that this effect does not obtain for speakers of French, a language which lacks vowel harmony. 
    This implies that Turkish speakers have internalized the tendency of harmonic sequences to be root-internal and of disharmonic transitions to cross word boundaries.

@@

To test for associations between the process (more specifically, the constraint that it imposes on roots) and type frequency in this database, each root was sorted into a $2 \times 2$ contingency table; the contents of each cell are specific to the process in question. The counts in this table are not expected to add up to 9,601, since many roots neither exemplify nor violate the process in question; for instance, monosyllabic words are irrelevant to root harmony. The Fisher exact test is used to compute a $p$-value representing the probability of the observed patterns arising under the null hypothesis that there is no association between the constraint and type frequency.

@@

\textsc{Backness Harmony} is exemplified in the lexicon insofar as there is a positive association between the backness of vowels in all adjacent syllables. Any disagreements on the backness specification of vowels in adjacent syllables are counted solely as exceptions, even if other vowel transitions in the root are harmonic; further assumptions are necessary to determine whether a root which has one disharmonic transition may in any sense ``obey'' harmony elsewhere.
%This follows from \emph{SPE} theory of lexical exceptionality (see also \citealt{Gouskova2012}), in which phonological exceptionality is a property of underlying representations, not individual segments. 
To construct the contingency table, roots are binned according to the backness specification of the first nucleus, and that of following nuclei. For example, the first two syllables of \emph{adalet} `justice' are harmonic, but it is coded as disharmonic because there is a \emph{a\ldots{}e} transition later in the word. The resulting counts are shown in Table \ref{bhs}. 61\% of roots conform to \textsc{Backness Harmony}, and the interaction between the backness of the first and of the subsequent vowels
%predicted by root-internal \textsc{Backness Harmony} 
is significant.

@@

\citet{Zimmer1969} administers two variants of the paired nonce word rating task. 
The first used 23 native adult speakers who were permitted to select either nonce word as more like Turkish, or to indicate `no preference'. 
For the purposes below, `no preference' results are ignored. 
The second experiment used 32 native adults, none of whom appeared in the preceding study, and used a forced binary choice.

### 2013-01-17 4.tex
Pre-"revisions", with lots of material in the TODO

Following \citet[ chap.~8]{Duanmu2009} and \citet[ chap.~3]{Hammond1999a}, 
a wordlist is generated using the English portion of the CELEX database \citep{CELEX}. 
Only words coded as ``monomorphemic'' are used, and all words coded as non-native are excluded.
This stringent criterion for inclusion eliminates many of the exceptions noted by \citeauthor{Duanmu2009} or \citeauthor{Hammond1999a} in their studies. For instance, nearly all the exceptions to \textsc{Obstruent Voice Assimilation} (see \S\ref{ova} below) noted by \citet[74]{Hammond1999a} are 
either complex words (e.g., \emph{jurisdiction}, \emph{madcap}, \emph{tadpole}, \emph{scapegoat}, \emph{magpie}) or non-native (e.g., \emph{vodka}, \emph{smorgasbord}).

@@

\begin{example}[Latinate verbs and small clauses] 
\label{harley}
\begin{tabular}{l l l l@{} l}
a. & {show him the painting} & \alt{} & * & {exhibit him the painting} \\
b. & {drink himself stupid}  & \alt{} & * & {imbibe himself stupid}    \\
%c. & {show it off}           & \alt{} & * & {exhibit it off}           \\
c. & {break it off}           & \alt{} & * & {terminate it off}        \\
\end{tabular}
\end{example}

@@

\textsc{Nasal Place Assimilation} (e.g., \citealt[65f.]{Borowsky1986}, \emph{SPE}:85, \citealt[62]{Halle1985a}) permits [ŋ] to be described as a pure allophone of /n/, and accounts for certain patterns in Latinate prefix allomorphy.

@@

\paragraph{Derived constraints} 
By hypothesis, \textsc{Obstruent Voice Assimilation}, \textsc{Nasal Place Assimilation}, and \textsc{Degemination} rule out a large number of possible clusters. 
In all, they target 316 out of 920 possible clusters (34\%), and only 11 (3\%) of these clusters has non-zero frequency in the sample.
Together, these three processes define a simple classifier in which a cluster is predicted to be unattested if and only if it would be neutralized by one of these processes. 
Unsurprisingly, this results in increased precision and a significant improvement in accuracy compared to the null baseline (sign test $p = 4.0$\e{-09}).

@@

Since this model has numerous experimenter-defined parameters, a close replication of \citeauthor{Hayes2008a}'s original paper is attempted: both their implementation and phonological feature specifications are used here.
Following \citet{HayesInPress}, dictionary entries are syllabified using the procedure described in \S\ref{syllabification} and a novel feature [$\pm$
\textsc{Coda}] is added to allow the model to distinguish coda and onset co
nsonants.
Also, following \citeauthor{Hayes2008a}, constraints are limited to those s
panning as many as three segments and an ``accuracy schedule'' of [$.001$, 
$.01$, $.1$, $.2$, $.3$] is used.
Since the maximum entropy model produces slightly different scores on each 
run, the worst-performing of 10 runs is reported here, following \citeauthor{Hayes2008a}.
This model has the poorest recall of any model; compared to the derived constraints baseline, the constraints induced by the maximum entropy model are narrower.
This is particularly clear regarding possible clusters with a nasal coda followed by a non-homorganic obstruent, like *[m.kl]: the vast majority of such clusters, which would be neutralized by \textsc{Nasal Place Assimilation}, are unattested, but many are erroneously assigned the highest possible score by the maximum entropy model.

>>>>

Following \citet[ chap.~8]{Duanmu2009} and \citet[ chap.~3]{Hammond1999a}, 
a wordlist is generated using the English portion of the CELEX database \citep{CELEX}. 
Only words marked in CELEX as ``monomorphemic'' are used, and all words 
labeled non-native are excluded.
This stringent criterion for inclusion eliminates many of the exceptions noted by \citeauthor{Duanmu2009} or \citeauthor{Hammond1999a} in their studies. For instance, nearly all the exceptions to \textsc{Obstruent Voice Assimilation} (see \S\ref{ova} below) noted by \citet[74]{Hammond1999a} are 
either complex words (e.g., \emph{jurisdiction}, \emph{madcap}, \emph{tadpole}, \emph{scapegoat}, \emph{magpie}) or non-native (e.g., \emph{vodka}, \emph{smorgasbord}).

@@

\begin{example}[Syntactic restrictions on Latinate verbs]
\label{harley}
\begin{tabular}{l l l l@{} l}
a. & {show him the painting} & \alt{} & * & {exhibit him the painting} \\
b. & {drink himself stupid}  & \alt{} & * & {imbibe himself stupid}    \\
%c. & {show it off}           & \alt{} & * & {exhibit it off}           \\
c. & {break it off}           & \alt{} & * & {terminate it off}        \\
\end{tabular}
\end{example}

@@

\textsc{Nasal Place Assimilation} (e.g., \citealt[65f.]{Borowsky1986}, \emph{SPE}:85, \citealt[62]{Halle1985a}) 
%\citealt[269]{Fudge1969}) 
permits [ŋ] to be described as a pure allophone of /n/, and accounts for certain patterns in Latinate prefix allomorphy.

@@

\paragraph{Derived constraints} 
By hypothesis, \textsc{Obstruent Voice Assimilation}, \textsc{Nasal Place Assimilation}, and \textsc{Degemination} rule out a large number of possible clusters. 
In all, they target 316 out of 920 possible clusters (34\%), and only 11 (3\%) of these clusters have non-zero frequency in the sample.
Together, these three processes define a simple classifier in which a cluster is predicted to be unattested if and only if it would be neutralized by one of these processes. 
Unsurprisingly, this results in increased precision and a significant improvement in accuracy compared to the null baseline (sign test $p = 4.0$\e{-09}).

@@

Since this model has numerous experimenter-defined parameters, a close replication of \citeauthor{Hayes2008a}'s original paper is attempted: both their implementation and phonological feature specifications are used here.
Following \citet{HayesInPress}, dictionary entries are syllabified using the procedure described in Appendix \ref{syllabification} and a novel feature [$\pm$
\textsc{Coda}] is added to allow the model to distinguish coda and onset co
nsonants.
Also, following \citeauthor{Hayes2008a}, constraints are limited to those s
panning as many as three segments and an ``accuracy schedule'' of [$.001$, 
$.01$, $.1$, $.2$, $.3$] is used.
Since the maximum entropy model produces slightly different scores on each 
run, the worst-performing of 10 runs is reported here, following \citeauthor{Hayes2008a}.
This model has the poorest recall of any model; compared to the derived constraints baseline, the constraints induced by the maximum entropy model are narrower.
This is particularly clear regarding possible clusters with a nasal coda followed by a non-homorganic obstruent, like *[m.kl]: the vast majority of such clusters, which would be neutralized by \textsc{Nasal Place Assimilation}, are unattested, but many are erroneously assigned the highest possible score by the maximum entropy model.
